Manuscript

### 2025-04-11 Manuscript_MD_ScoreReLiability.qmd
update manuscript, init refs

---
title: "Manuscript"
format: html
editor: 
  markdown: 
    wrap: 72
---

@@

Large-scale collaborative replication efforts have sparked discussions
surrounding the replicability of psychological phenomena. For attempts
to estimate a phenomenon’s replicability or to predict whether a future
replication will replicate successfully, heterogeneity is a crucial
parameter that needs to be assessed. In the meta-analytic context,
heterogeneity describes the variation of effect sizes, free of sampling
error. Therefore, the presence of heterogeneity can imply that some
replications of a single phenomenon may be successful, while others are
not. If heterogeneity increases, meaning the phenomenon’s effect size
varies more strongly for unexplained reasons, the probability of
observing an effect size around zero or even in the negative space grows
larger as well. If we know the mean size and heterogeneity of a
phenomenon’s effect, we would theoretically be able to establish an
expected replication rate (reference).

Similarly, it has been argued that heterogeneity in effect sizes is an
indicator of the theory’s “completeness” surrounding the phenomenon.
Linden and Hönekopp argue that “low (as opposed to high) heterogeneity
reflects a more advanced understanding of the subject matter being
studied” (2021, p.2). Similarly, Schuetze and von Hippel (2023) argue
that heterogeneity in effects is an indicator of a vague, poorly
specified theory.

Large-scale attempts of direct replications, using identical protocols,
such as the Many Labs studies or Registered Replication Reports
(references), for the first time allow researchers to estimate
heterogeneity undistorted by selection processes like publication bias.
In re-analyses of these studies, Olsson-Collentine et al. (2020)
identify a positive correlation between a phenomenon’s effect size and
its heterogeneity. Similarly, van Erp et al. (2017) and Stanley et al.
(2018) estimate strong degrees of heterogeneity across conceptual
replications in Psychology. In a separate re-analysis of large scale
direct replications, Renkewitz et al. (in preparation) identify
substantial heterogeneity in almost all projects where a non-zero effect
could be identified. This aligns with the correlation found by
Olsson-Collentine et al. (2020).

One source for this heterogeneity, which is not a typical moderator tied
to a specific theory, might be impairments in score reliability which
vary in their extent across replications. This is an argument put
forward repeatedly by original researchers and replicators alike (refs).
Hunter and Schmidt (xxx) and Wiernik and Dahlke (xxx), discussing the
role of score reliability for standardized effect sizes, claim that
differences in score reliability should inflate, and therefore increase,
heterogeneity. This implies that standard corrections for imperfect
score reliability, such as attenuation correction procedures, should
remove that inflation, reducing heterogeneity in corrected effect sizes.
Over the following pages, we assess whether score reliability can in
fact be made responsible for heterogeneity identified across phenomena.
We find no empirical evidence for the argument that correcting for
differences in score reliability reduces effect size heterogeneity.
Instead, to our great surprise, we find that heterogeneity might as well
be larger after the correction procedure. Therefore, theoretical
analyses regarding the relationship between score reliability and
heterogeneity need to be false. In the second half of this text we make
use of analytical arguments to identify the necessary conditions under
which an attenuation correction inevitably leads to an increase in
effect size heterogeneity.

@@

In both the initial reports of large scale replication attempts, as well
as the re-analyses by Olsson-Collentine et al. (2020), standardized
effect sizes (from here-on abbreviated as ES) such as Cohen’s d or
Hedge’s g were used. For the remainder of the article, Cohen’s d,
defined in @eq-d, will be used as an exemplary estimate of ES, as it is
used across a wide range of contexts and well understood by a broad
audience.

@@

Here, MD refers to the mean difference between two groups of interest,
while $\sigma_X$ is the total pooled standard deviation. Since $d$
constitutes an observed effect size it is affected by sampling error. In
terms of heterogeneity, we are not interested in the variance of $d$,
but the variance of the true, underlying ES $\delta$. In a
random-effects meta-analytic model, we can use the standard error and an
estimate of heterogeneity to understand how far off the individual
observed $d$ is from the true $\delta$ in a weighting procedure
(reference). The standard error of Cohen’s $d$ is computed as defined in
@eq-SEd.

@@

As is widely known, score reliability affects ES. In the context of
classical test theory (CTT) score reliability $\rho_{XX’}$ is defined as
the ratio of true to total score variance, as defined in @eq-rel.

@@

In this equation $\sigma_T^2$ refers to the true score variance in the
sense of CTT, meaning the actual variance of the variable, undistorted
by random measurement error. In the same sense, $\sigma_X^2$ refers to
the total variance of scores, including both the true variance and the
random error variance $\sigma_E^2$. This implies that, if true score
variance is assumed to be constant across replications, a lower score
reliability can only occur due to a larger error score variance. Total
score variance would also be larger, therein leading to a smaller ES, as
opposed to a similar data-set where a higher score reliability is
achieved through a smaller error score variance. However, score
reliability is an aspect of a measuring instrument applied to a
population. In the meta-analytic context, as studies tend to be
replicated in different populations, neither true nor error score
variance can be expected to always be identical across replications.
This most likely leads to heterogeneity in score reliability, which, as
discussed, is bound to affect ES heterogeneity as well.

Previous discussions of heterogeneity in score reliability have
exclusively discussed it as a parameter that inflates heterogeneity in
ES, implying that, if score reliability was perfect across all
replications, the actual heterogeneity would have been lower. In their
discussion, Wiernik and Dahlke claim that „*Measurement error variance
will impact the results of meta-analyses in three ways: by (a) biasing
the mean effect size toward zero, (b) inflating effect-size
heterogeneity and confounding moderator effects, and (c) confounding
publication-bias and sensitivity analyses*” (p. 3, 2020). Additionally,
more clearly, they state that “*If the studies included in a
meta-analysis differ in their measures’ reliabilities, heterogeneity
estimates will be artifactually inflated, erroneously suggesting larger
potential moderator effects*” (p. 4, Wiernik & Dahlke, 2020). They base
their claims largely on work done by Hunter and Schmidt (2014), who
claim that “*Variation in reliability across studies causes variation in
the observed effect sizes above and beyond that produced by sampling
error.*” (p. 302). Overall, both references imply that differences in
score (un)reliability inevitably lead to an inflated heterogeneity in
ES.

As both sources Hunter and Schmidt (2014) and Wiernik and Dahlke (2020)
will be referred to repeatedly, we abbreviate these references as H&S
and W&D respectively.

@@

If information concerning score reliability is available, it is possible
to correct the individual ES for its unreliability. This process is also
known as attenuation correction and, while not without its criticisms
(reference), is a widespread practice in Psychology (reference). @eq-dc
describes a simple attenuation correction procedure.

@@

Here, $d_c$ describes the attenuation-corrected estimate of $d$, which
is corrected by dividing it by the estimate of score reliability
$\hat{\rho}_{xx’}$. As described in W&D or Lord & Novick (references),
this is essentially the same, as if the corrected estimate of $d_c$ was
constructed using an estimate of true score standard deviation, as in
@eq-dct.

@@

Re-analyses of the collaborative multi-site replication efforts have
uncovered differences in score reliability across replications of the
same phenomena (ref McShaw etc.). The claims made in W&D and H&S imply
that the ES heterogeneity uncovered in these projects therefore could
potentially be explained by these differences in score reliability.
Similarly, attempting to explain the substantial extent of ES
heterogeneity identified in psychological phenomena, both the original
authors as well as the multi-site replication authors claim that
differences in score reliability might be responsible (references).

@@

We have collected a large number of openly available multi-site direct
replications on psychological phenomena (Fünderich et al. 2024). The
data-sets of about 50 phenomena, largely stemming from the efforts of
the Many Labs studies, Registered Replication Reports or the
Psychological Science Accelerator, are made openly available in a
standardised format at \[osf.io\], the DRIPHT repository.

@@

Concerning the re-arranged inequality from Appendix A, we identify one
crucial condition or rule described in @eq-ineq_short.

@@

As long as $R_1^3$ is smaller or equal compared to $R_2$, @eq-ineq_init
is bound to hold true. This means that as long as the average score
reliability, to the power of 3, does not exceed the relative remaining
variance heterogeneity after the attenuation correction, @eq-ineq_init
is not violated. For example, assuming the average score reliability is
about .8, as long as $R_2$ is at least .512 or larger, the inequality
holds true and ES heterogeneity grows larger as a consequence of the
attenuation correction procedure. In such cases, the claims made in W&D
and H&S could not hold.

$R_1$ essentially describes the relative size of mean score variance
after the attenuation correction, and $R_2$ essentially described the
relative size of score variance heterogeneity after the attenuation
correction. Therefore, @eq-ineq_init states that for such a correction
procedure to truly reduce ES heterogeneity, the reduction in the
heterogeneity of score variance needs to be substantially larger than
the reduction in its mean values. The mean value reduction carries
stronger weight, as $R_1$ is taken to the power of 3 in @eq-ineq_init.
The smaller $R_1$ or $R_2$, the larger the reduction in heterogeneity or
mean value.

As long as there are substantial differences in the true score variance
across administrations, it is unlikely that an attenuation correction
can actually produce such a low $R_2$. Generally speaking, $R_1$, and
therefore the reduction in mean score variance, largely depends on the
relative measurement precision. If measurements were made more
precisely, average score reliability and $R_1$ would be higher. On the
other hand $R_2$, and therefore the reduction in heterogeneity of score
variance, depends on how different the measurement precision was at the
individual administrations, but also how different the individual
administrations sites are regarding their true score variance. If the
administration sites carry large differences in true score variance
$R_2$ is unlikely to be substantially smaller than $R_1$.

It is important to note that the condition in @eq-ineq_short only
describes under which conditions the inequality is guaranteed to hold,
no matter what value other parameters like the mean value or
heterogeneity in mean differences take on. This means that, even if the
condition in @eq-ineq_short is violated, it is by no means guaranteed
that an attenuation correction procedure would actually reduce
heterogeneity in ES. This means that inference made from @eq-ineq_short
is asymmetrical. While we know, if the inequality holds, heterogeneity
in corrected ES will be larger than heterogeneity in uncorrected ES, we
can't be certain that the opposite is true if the inequality does not
hold.

***\[maybe move to discussion further below\]*** The majority of studies
in the social sciences make use of convenience sampling. Often,
respondents are collected from a local student pool, where students
themselves would choose which studies to partake in. In such cases, we
believe there is little reason for us to expect homogeneity in true
score variance across administration sites. As participants across
administrations self-select into the study, no real sampling process
takes place - implying that we have no reason to assume that the true
score variance is stable across these administration sites. However,
since researchers are well aware of the importance of measurement
precision, which is also held stable more easily, we believe there is
little reason for $R_1$ to be extremely low.

@@

@tbl-tauES showed that for all phenomena, where we identified
statistically significant ES heterogeneity, the heterogeneity was even
larger after the attenuation correction procedure. Subsequently, in
Appendix A we demonstrate that if the $R_2$ metric is larger than
$R_1^3$, ES heterogeneity is bound to be larger after applying said
procedure. Here, we will re-analyse the data, estimating $R_1$ and $R_2$
for each phenomenon respectively. This allows us to assess in how far
the estimates of ES heterogeneity align with our claims regarding the
$R_1$ and $R_2$ metrics.

@@

Estimation of $R_1$ and $R_2$ require estimates of meta-analytic mean
and heterogeneity of both the total score variance $\sigma^2_X$ and the
true score variance $\sigma^2_T$. Elsewhere, we proposed the method
boot-err (ref.) as an alternative method to test for and estimate the
differences in error score variance $\sigma^2_E$. In Beinhauer et al.
(ref.), we deliver detailed instructions on how estimates for mean and
heterogeneity of total score variance, and the score variance components
can be derived. For the total score variance $\sigma^2_X$, all we need
to do is to appropriately transform the observed estimates using the
variance-stabilizing properties of the natural log. Subsequently, we use
bootstrapping to derive robust standard errors for these transformed
observations. Using a random-effects meta-analysis, we derive estimates
of mean and heterogeneity, which need to be backtransformed to the
original scale. Appropriate equation to back-transform estimates can be
found in Appendix B.

For the true score variance $\sigma^2_T$, we found that directly
estimating its heterogeneity, even if an appropriate
variance-stabilizing transformation is used, is severely biased.
Instead, we can estimate the error score variance $\sigma^2_E$ using an
estimate of score reliability. Performing the boot-err, which, again,
implies transformations using the natural log, deriving standard errors
using Bootstrapping, running a random-effects meta-analysis and
back-transforming the estimates, nets us estimates of mean and
heterogeneity of $\sigma^2_E$. Since we already have the same estimates
available for the total score variance, we can estimate meta-analytic
mean and heterogeneity of true score variance.

$$\hat{\mu}_{\sigma^2_T} = \hat{\mu}_{\sigma^2_X} - \hat{\mu}_{\sigma^2_E} \ \ \ \ \ \ and \ \ \ \ \  \hat{\tau}^2_{\sigma^2_T} = \hat{\tau}^2_{\sigma^2_X} - \hat{\tau}^2_{\sigma^2_E}$$ {#eq-meantauT}

@eq-meantauT provides estimates of meta-analytic mean of true score
variance and heterogeneity. These estimates are largely unbiased, unlike
estimates stemming from using the boot-err to directly run meta-analytic
models on true score variance estimates. Interested readers are referred
to Beinhauer et al. (ref.)

@@

For the two exceptional phenomena where the attenuation correction
procedure actually led to an increase in ES heterogeneity
(Srull_Behavior_Hostility and Shnabel_Willingness_Reconcile_RPP), the
significance test for the ES heterogeneity was not significant, both
before and after the correction. The results regarding $R_2$ indicate
that the majority of score variance heterogeneity was due to differences
in true score variance $\sigma^2_T$. Since either the heterogeneity was
so small or the sample size so low that the heterogeneity could not be
statistically distinguished from zero, we should be cautious in the
interpretation of a reduction in that estimate. Most likely, the
reduction in heterogeneity is better explained by imprecise estimation
than an actual explanation of unexplained heterogeneity.

@@

Generally, @tbl-tauES implies that for almost all measurements
concerning these phenomena, the heterogeneity in observed score
variances could not be substantially explained by differences in
measurement quality. Only for one phenomenon, where differences in score
variance could be found, $R_2$ was smaller than $R_1^3$. As discussed,
for this phenomenon, the number of administration sites was too low to
accurately estimate heterogeneity in score variance components. For all
remaining phenomena, $R_2$ was larger than $R_1$, implying that an
attenuation correction procedure could not reduce ES heterogeneity. The
results in @tbl-tauES agree with this conclusion, as no phenomena could
be identified where statistically significant ES heterogeneity was
accompanied by a reduction in differences due to the correction.

@@

We have examined meta-analytic heterogeneity in standardized mean
difference. Since score reliability is often discussed as a source for
such heterogeneity, we estimated the heterogeneity before and after
correcting for differences in reliability coefficients using an
attenuation correction procedure. We found that a) in the majority of
cases, the change in ES heterogeneity is small to non-existent. This
implies that, across the data-sets we got to examine, differences in
score reliability do not serve as an appropriate explanation for ES
heterogeneity. We also found that b) in the majority of cases, the
direction of change in ES heterogeneity does not align with assumptions
formulated in W&D and H&S. In almost all cases (every single case, if we
restrict ourself to phenomena where we found statistically significant
differences in ES), the attenuation correction led to increased
estimates of ES heterogeneity. This implies that the claims found in W&D
and H&S stem from a general misunderstanding of how score reliability
can affect ESs.

Using analytical arguments, we derived a broad rule, outlining under
which conditions we can expect an attenuation correction to reduce ES
heterogeneity. In general, score reliability affects ESs by inflating
the score variance, as the (square root of the) score variance is used
to standardize the mean difference. The conditions we found depend on
how both the mean value and the heterogeneity of score variance are
reduced by correcting for imperfect score reliability. We introduced two
novel metrics: $R_1$, describing the fraction of mean true score
variance in the mean total score variance, therefore mirroring the
average score reliability coefficient, and $R_2$, describing the
fraction of true score variance heterogeneity in the total score
variance heterogeneity. We find that, as long as $R_1^3 < R_2$, an
attenuation correction will increase ES heterogeneity.

Since $R_1$ is carrying the power of 3, it implies that the reduction in
mean score variance is weighted more heavily than the reduction in score
variance heterogeneity. This is best described through an example: We
know what an acceptable average score reliability looks like in the area
of Psychology. Researchers largely target a score reliability of at
least .7, preferably .8. An average score reliability of .7 implies that
70% of the variance observed in scores can actually be attributed to
true underlying differences. From our derived rule, we know that, as
long as $R_2$ is larger or equal to $.7^3 = .343$, the attenuation
correction procedure can only increase ES heterogeneity. If we want to
explain ES heterogeneity through score reliability, even though only 30%
of the variance observed in score can be attributed to random
measurement error, at least 65.7%, almost two thirds, of the
heterogeneity in score variance would need to be attributable to
differences in the error score variance. In other words, at most 34.3%
of heterogeneity in score variances may be due to actual differences in
how the latent variable is distributed (in terms of spread) across the
administration sites.

This also means that ES heterogeneity is best reduced, if average score
reliability ($R_1$) is substantially large. In such a scenario, the
$R_2$ is more likely to be low enough for the inequality to no longer
hold. The larger the average score reliability, the more likely it is
that the attenuation correction procedure can actually reduce ES
heterogeneity. However, as the correction would only be rather small in
such cases, it remains to be seen whether the reduction in ES
heterogeneity would be sufficient to claim that differences in score
reliability actually posed a relevant moderator in such a case.

@@

Taken together, these facets lead to what we observed throughout this
text: Reliability coefficients are in an acceptable range, leading to a
small, but constant reduction in mean score variance if we apply an
attenuation correction. At the same time, there is substantial variation
in true score variance, implying that most of the score variance
heterogeneity can not be attributed to differences in error score
variance. Since only differences in error score variance can be
corrected for, the attenuation correction has a very small effect on
heterogeneity estimates in score variance. Therefore, as indicated by
our rule, an attenuation correction in these setting can rarely deliver
a reduction in ES heterogeneity. This also means that, in such settings,
score reliability does not explain any ES heterogeneity. If anything, it
masks it to a small degree.

If we, in the space of Psychology, keep making use of convenience
samples simply because they are so convenient, we can not expect
anything to change. Most likely, ES heterogeneity will remain a
substantial factor hampering the interpretability of original and
replication studies. We could learn more about standardized effect sizes
and why our phenomena vary across settings, populations or continents by
not following the current status quo of convenience samples. Until this
status quo changes, however, we will keep wasting valuable resources by
attempts of discussing ES heterogeneity or assessing uninformative
differences in score reliability.

These results fit in with work recently published/preprinted by
Olsson-Collentine et al. (2023). In a large simulation scheme, they
demonstrate that differences in score reliability across administrations
typically deflate heterogeneity in uncorrected correlations. Only as the
true heterogeneity in correlations grows larger, score reliability
differences actually inflate heterogeneity as claimed in W&D. While
correlations and standardized effect sizes are not identical, the way
score (un)reliability affects these parameters is highly similar.

>>>>

---
title: "Manuscript"
bibliography: ReLiability_references.bib
csl: styles/apa-6th-edition.csl
format: html
editor: 
  markdown: 
    wrap: 72
---

@@

Large-scale collaborative replication efforts have sparked discussions
surrounding the replicability of psychological phenomena. For attempts
to estimate a phenomenon’s replicability or to predict whether a future
replication will replicate successfully, heterogeneity is a crucial
parameter that needs to be assessed [@stanley2018]. In the meta-analytic
context, heterogeneity describes the variation of effect sizes, free of
sampling error [@dersimonian1986; @higgins2002; @riley2011]. Therefore,
the presence of heterogeneity can imply that some replications of a
single phenomenon may be successful, while others are not. If
heterogeneity increases, meaning the phenomenon’s effect size varies
more strongly for unexplained reasons, the probability of observing an
effect size around zero or even in the negative space grows larger as
well [@kenny2019]. If we know the mean size and heterogeneity of a
phenomenon’s effect, we would theoretically be able to establish an
expected replication rate [e.g. @vu2024].

Similarly, it has been argued that heterogeneity in effect sizes is an
indicator of the theory’s “completeness” surrounding the phenomenon.
Linden and Hönekopp argue that “low (as opposed to high) heterogeneity
reflects a more advanced understanding of the subject matter being
studied” [-@LindenHönekopp, p. 2]. Similarly, von Hippel and Schuetze
[-@vonHippel2023] argue that heterogeneity in effects is an indicator of
a vague, poorly specified theory.

Large-scale attempts of direct replications, using identical protocols,
such as the Many Labs studies or Registered Replication Reports [e.g.
@ManyLabs1; @RRHart], for the first time allow researchers to estimate
heterogeneity undistorted by selection processes like publication bias.
In re-analyses of these studies, Olsson-Collentine et al. [-@olsson2020]
identify a positive correlation between a phenomenon’s effect size and
its heterogeneity. Similarly, van Erp et al. [-@vanErp2017] and Stanley
et al. [-@stanley2018] estimate strong degrees of heterogeneity across
conceptual replications in Psychology. In a separate re-analysis of
large scale direct replications, Renkewitz et al. (in preparation)
identify substantial heterogeneity in almost all projects where a
non-zero effect could be identified. This aligns with the correlation
found by Olsson-Collentine et al. [-@olsson2020].

One source for this heterogeneity, which is not a typical moderator tied
to a specific theory, might be impairments in score reliability which
vary in their extent across replications. \[This is an argument put
forward repeatedly by original researchers and replicators alike (do we
actually know of any?).\] Hunter and Schmidt [-@hunter2004] and Wiernik
and Dahlke [-@wiernik2020], discussing the role of score reliability for
standardized effect sizes, claim that differences in score reliability
should inflate, and therefore increase, heterogeneity. This implies that
standard corrections for imperfect score reliability, such as
attenuation correction procedures, should remove that inflation,
reducing heterogeneity in corrected effect sizes. Over the following
pages, we assess whether score reliability can in fact be made
responsible for heterogeneity identified across phenomena. We find no
empirical evidence for the argument that correcting for differences in
score reliability reduces effect size heterogeneity. Instead, to our
great surprise, we find that heterogeneity might as well be larger after
the correction procedure. Therefore, theoretical analyses regarding the
relationship between score reliability and heterogeneity need to be
false. In the second half of this text we make use of analytical
arguments to identify the necessary conditions under which an
attenuation correction inevitably leads to an increase in effect size
heterogeneity.

@@

In both the initial reports of large scale replication attempts, as well
as the re-analyses by Olsson-Collentine et al. [-@olsson2020],
standardized effect sizes (from here-on abbreviated as ES) such as
Cohen’s d or Hedge’s g were used. For the remainder of the article,
Cohen’s d, defined in @eq-d, will be used as an exemplary estimate of
ES, as it is used across a wide range of contexts and well understood by
a broad audience.

@@

Here, MD refers to the mean difference between two groups of interest,
while $\sigma_X$ is the total pooled standard deviation. Since $d$
constitutes an observed effect size it is affected by sampling error. In
terms of heterogeneity, we are not interested in the variance of $d$,
but the variance of the true, underlying ES $\delta$. In a
random-effects meta-analytic model, we can use the standard error and an
estimate of heterogeneity to understand how far off the individual
observed $d$ is from the true $\delta$ in a weighting procedure
[@borenstein2009] The standard error of Cohen’s $d$ is computed as
defined in @eq-SEd.

@@

As is widely known, score reliability affects ES [e.g. @hunter2004;
@wiernik2020]. In the context of classical test theory (CTT) score
reliability $\rho_{XX’}$ is defined as the ratio of true to total score
variance [@lord2008], as defined in @eq-rel.

@@

In this equation $\sigma_T^2$ refers to the true score variance in the
sense of CTT, meaning the actual variance of the variable, undistorted
by random measurement error. In the same sense, $\sigma_X^2$ refers to
the total variance of scores, including both the true variance and the
random error variance $\sigma_E^2$. This implies that, if true score
variance is assumed to be constant across replications, a lower score
reliability can only occur due to a larger error score variance. Total
score variance would also be larger, therein leading to a smaller ES, as
opposed to a similar data-set where a higher score reliability is
achieved through a smaller error score variance. However, score
reliability is an aspect of a measuring instrument applied to a
population [@lord2008; @vacha1998]. In the meta-analytic context, as
studies tend to be replicated in a variety of populations, neither true
nor error score variance can be expected to always be identical across
replications. This most likely leads to heterogeneity in score
reliability, which, as discussed, is bound to affect ES heterogeneity as
well.

Previous discussions of heterogeneity in score reliability have
exclusively discussed it as a parameter that inflates heterogeneity in
ES, implying that, if score reliability was perfect across all
replications, the actual heterogeneity would have been lower. In their
discussion, Wiernik and Dahlke claim that „*Measurement error variance
will impact the results of meta-analyses in three ways: by (a) biasing
the mean effect size toward zero, (b) inflating effect-size
heterogeneity and confounding moderator effects, and (c) confounding
publication-bias and sensitivity analyses*” [p. 3, -@wiernik2020].
Additionally, more clearly, they state that “*If the studies included in
a meta-analysis differ in their measures’ reliabilities, heterogeneity
estimates will be artifactually inflated, erroneously suggesting larger
potential moderator effects*” [p. 4, -@wiernik2020]. They base their
claims largely on work done by Hunter and Schmidt [-@hunter2004], who
claim that “*Variation in reliability across studies causes variation in
the observed effect sizes above and beyond that produced by sampling
error.*” (p. 302). Overall, both references imply that differences in
score (un)reliability inevitably lead to an inflated heterogeneity in
ES.

As both sources Hunter and Schmidt [-@hunter2004] and Wiernik and Dahlke
[-@wiernik2020] will be referred to repeatedly, we abbreviate these
references as H&S and W&D respectively.

@@

If information concerning score reliability is available, it is possible
to correct the individual ES for its unreliability. This process is also
known as attenuation correction and, while not without its criticisms
[@muchinsky1996; @schmidt1996; @schennach2016], is a widespread practice
in Psychology (reference). @eq-dc describes a simple attenuation
correction procedure.

@@

Here, $d_c$ describes the attenuation-corrected estimate of $d$, which
is corrected by dividing it by the estimate of score reliability
$\hat{\rho}_{xx’}$. As described in W&D or Lord & Novick [-@lord2008],
this is essentially the same, as if the corrected estimate of $d_c$ was
constructed using an estimate of true score standard deviation, as in
@eq-dct.

@@

Re-analyses of the collaborative multi-site replication efforts have
uncovered differences in score reliability across replications of the
same phenomena [@shaw2020]. The claims made in W&D and H&S imply that
the ES heterogeneity uncovered in these projects therefore could
potentially be explained by these differences in score reliability.
Similarly, attempting to explain the substantial extent of ES
heterogeneity identified in psychological phenomena, both the original
authors as well as the multi-site replication authors claim that
differences in score reliability might be responsible (again, do we
actually know of any?).

@@

We have collected a large number of openly available multi-site direct
replications on psychological phenomena [@dripht2024]. The data-sets of
about 50 phenomena, largely stemming from the efforts of the Many Labs
studies, Registered Replication Reports or the Psychological Science
Accelerator, are made openly available in a standardised format at
\[<https://osf.io/7p8rk/>\], the DRIPHT repository.

@@

Concerning the re-arranged inequality from Appendix A, we identify one
necessary but not sufficient condition or rule described in
@eq-ineq_short.

@@

With necessary but not sufficient we mean that, as long as
@eq-ineq_short holds, an attenuation correction will inevitably increase
ES heterogeneity. However, if the inequality does not hold, it is not
guaranteed that the attenuation correction can decrease ES
heterogeneity. Whether ES heterogeneity can be decreased depends on
other parameters, that we left out of this short form, but can be found,
accompanied by a more detailed description, in @eq-ineq_fin.

As long as $R_1^3$ is smaller or equal compared to $R_2$, @eq-ineq_init
is bound to hold true. This means that as long as the average score
reliability, to the power of 3, does not exceed the relative remaining
variance heterogeneity after the attenuation correction, @eq-ineq_init
is not violated. For example, assuming average score reliability is
about .8, the heterogeneity in $\sigma^2_T$ needs to make up less than
51.2% ($R_2 \geq .8^3 = .512$) of the heterogeneity in total score
variance $\sigma^2_X$. At $R_1 = .8$, as long as $R_2 \geq .512$, the
claims made in W&D and H&S can not hold.

$R_1$ essentially describes the relative size of mean score variance
after the attenuation correction, and $R_2$ essentially described the
relative size of score variance heterogeneity after the attenuation
correction. Therefore, @eq-ineq_short states the following regarding the
change in mean value and heterogeneity due to the correction: ES
heterogeneity will always be reduced, as long as the remaining
heterogeneity in score variance is not substantially smaller than the
size of the remaining mean value.

As long as there are substantial differences in the true score variance
across administrations, it is unlikely that an attenuation correction
can actually produce such a low $R_2$. Generally speaking, $R_1$ largely
depends on the relative measurement precision. If measurements were made
more precisely, average score reliability and $R_1$ would be higher. On
the other hand, $R_2$ depends on how different the measurement precision
was at the individual administrations, but also how different the
individual administrations sites are regarding their true score
variance. If the administration sites carry large differences in true
score variance $R_2$ is unlikely to be substantially smaller than $R_1$.

It is important to note that the condition in @eq-ineq_short only
describes under which conditions the inequality is guaranteed to hold,
no matter what value other parameters like the mean value of $MD$ or its
heterogeneity take on. This means that, even if the condition in
@eq-ineq_short is violated, it is by no means guaranteed that an
attenuation correction procedure would actually reduce heterogeneity in
ES. This means that inference made from @eq-ineq_short is asymmetrical.
While we know, if the inequality holds, heterogeneity in corrected ES
will be larger than heterogeneity in uncorrected ES, we can't be certain
that the opposite is true if the inequality does not hold.

@@

@tbl-tauES showed that for all phenomena, for which we identified
statistically significant ES heterogeneity, the heterogeneity was even
larger after the attenuation correction procedure. Here, we will
re-analyse the data, estimating $R_1$ and $R_2$ for each phenomenon
respectively. Therein, we not only show that our claims derived from
analytical arguments hold up in empirical assessment, but also
demonstrate how the metrics $R_1$ and $R_2$ may be used to inform how
score reliability affected ES heterogeneity.

@@

Estimation of $R_1$ and $R_2$ requires estimates of meta-analytic mean
and heterogeneity of both the total score variance $\sigma^2_X$ and the
true score variance $\sigma^2_T$. Elsewhere, we proposed the method
Boot-Err (ref.) as an alternative method to test for and estimate the
differences in error score variance $\sigma^2_E$. Among other things, in
Beinhauer et al. (ref.), we deliver detailed instructions on how
estimates for mean and heterogeneity of total score variance, and the
score variance components can be derived. This means that Boot-Err can
be used to generate all estimates required for computation of the
metrics $R_1$ and $R_2$.

For the following analysis, Boot-Err is used. As Boot-Err makes use of
meta-analytic techniques and bootstrap sampling, we perform analyses
using the packages *metafor*, version 4.8-0, and *boot*, version 1.3-31,
in the statistical programming language *R*, version 4.4.2.

@@

For the two exceptions, those phenomena where the attenuation correction
procedure actually led to an decrease in ES heterogeneity
(Srull_Behavior_Hostility and Shnabel_Willingness_Reconcile_RPP), the
significance test for the ES heterogeneity was not significant, both
before and after the correction. The results regarding $R_2$ indicate
that the majority of score variance heterogeneity was due to differences
in true score variance $\sigma^2_T$. Since either the heterogeneity was
so small or the sample size so low that the heterogeneity could not be
statistically distinguished from zero, we should be cautious in the
interpretation of a reduction in that estimate. Most likely, the
reduction in heterogeneity is better explained by imprecise estimation
than an actual reduction of heterogeneity.

@@

Generally, @tbl-R1R2 implies that for all measurements concerning these
phenomena, the heterogeneity in observed score variances could not be
substantially explained by differences in measurement quality. For all
phenomena, $R_2$ was larger than $R_1$, implying that an attenuation
correction procedure could not reduce ES heterogeneity in any of these
cases. The results in @tbl-tauES agree with this conclusion, as no
phenomena could be identified where statistically significant ES
heterogeneity was accompanied by a reduction in differences due to the
correction.

@@

We have examined meta-analytic heterogeneity in standardized mean
differences. Since score reliability is often discussed as a source for
such heterogeneity, we estimated the heterogeneity before and after
correcting for differences in reliability coefficients using an
attenuation correction procedure. We found that a) in the majority of
cases, the change in ES heterogeneity is small to non-existent. This
implies that, across the data-sets we got to examine, differences in
score reliability do not serve as an appropriate explanation for ES
heterogeneity. We also found that b) in the majority of cases, the
direction of change in ES heterogeneity does not align with assumptions
formulated in W&D and H&S. In almost all cases (every single case, if we
restrict ourself to phenomena where we found statistically significant
differences in ES), the attenuation correction led to increased
estimates of ES heterogeneity. This means that the claims found in W&D
and H&S, implying that an attenuation correction can reduce ES
heterogeneity, stem from a general misunderstanding of how score
reliability can affect ESs.

Using analytical arguments, we derived a broad rule, outlining under
which conditions we can expect an attenuation correction to increase ES
heterogeneity. In general, score reliability affects ESs by inflating
the score variance, as the (square root of the) score variance is used
to standardize the mean difference. The conditions we found depend on
how both the mean value and the heterogeneity of score variance are
reduced by correcting for imperfect score reliability. We introduced two
novel metrics: $R_1$, describing the fraction of mean true score
variance in the mean total score variance, therefore mirroring the
average score reliability coefficient, and $R_2$, describing the
fraction of true score variance heterogeneity in the total score
variance heterogeneity. We find that, as long as $R_1^3 < R_2$, an
attenuation correction will increase ES heterogeneity.

Since $R_1$ is carrying the power of 3, it implies that the reduction in
mean score variance is affecting the ES heterogeneity more strongly than
the reduction in score variance heterogeneity. This is best described
through an example: We know what an acceptable average score reliability
looks like in the area of Psychology. Researchers largely target a score
reliability of at least .7, preferably .8. An average score reliability
of .7 implies that 70% of the variance observed in scores can actually
be attributed to true underlying differences. From our derived rule, we
know that, as long as $R_2$ is larger or equal to $.7^3 = .343$, the
attenuation correction procedure can only increase ES heterogeneity. If
we want to explain ES heterogeneity through score reliability, even
though only 30% of the variance observed in score can be attributed to
random measurement error, at least 65.7%, almost two thirds, of the
heterogeneity in score variance would need to be attributable to
differences in the error score variance. In other words, at most 34.3%
of heterogeneity in score variances may be due to actual differences in
how the latent variable is distributed (in terms of spread) across the
administration sites.

@@

The combination of these aspects can explain what we observed throughout
this text: Reliability coefficients are in an acceptable range, leading
to a small, but constant reduction in mean score variance if we apply an
attenuation correction. At the same time, there is substantial variation
in true score variance, implying that most of the score variance
heterogeneity can not be attributed to differences in error score
variance. Since only differences in error score variance can be
corrected for, the attenuation correction has a very small effect on
heterogeneity estimates in score variance. Therefore, as indicated by
our inequality, an attenuation correction is likely to have very little
impact on ES heterogeneity. If anything, it most likely increases it.
This also means that, in such settings, score reliability does not serve
as a moderator explaining ES heterogeneity.

If we, in the field of Psychology and other social or behavioural
sciences, keep making use of convenience samples simply because they are
so convenient, we can not expect anything to change. Most likely, ES
heterogeneity will remain a substantial factor hampering the
interpretability of original and replication studies. Instead, we should
focus our data collection efforts on the generation of samples, where we
know that the variance in the sample means something. This is most
easily achieved by proper sampling techniques, which ensure that the
sample is representative for some population. In such cases, it might be
that the true score variance varies less strongly than what we currently
observe. And should the heterogeneity in true score variance remain as
large, than we know that this is meaningful information from which we
can draw meaningful inferences. Therein, we could learn more about
standardized effect sizes and why our phenomena vary across settings,
populations or continents by not following the current status quo of
convenience samples. Until this status quo changes, however, we will
keep wasting valuable resources by attempts of discussing ES
heterogeneity or assessing uninformative differences in score
reliability.
